Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-05 paper/paper.tex
minor revisions

\begin{itemize}
  \item The components of the configuration are lifted as arguments of the evaluation
    function.
  \item The refunctionalized AAM collects all possible configurations, but our final target is an 
    abstract definitional interpreter that cares about final evaluation results.
    So starting from this section our abstract semantic artifact returns a set of final 
    values instead of collected configurations. 
  \item Configurations still play an important role in caching, where a cache
    is a mapping from configurations to sets of final values. The caching algorithm
    ensures termination of the analysis, as well as performs a sound over-approximation
    of all possible values and reachable paths.
  \item We simplify the continuations with a nondeterministic operator @nd@ such 
    that the function @aeval@ only uses one layer of continuation.
\end{itemize}

@@

We first transform the function @aeval@ to direct-style by removing the additional 
continuation argument @continue@ and sequentializing the expressions in @aeval@.
To enable using delimited control operators inside of @aeval@ we also add an 
annotation \verb|@cps[Ans]| on the return type, which tells the compiler to CPS
transform this function.
Back to direct-style on the case where @e@ appears as an atomic expression is straightforward: 
we simply remove the application of @continue@. Now the abstract evaluator @aeval@ looks as follows:

@@

The case where @e@ is a @let@ expression is not that obvious. As previously identified, 
function @nd@ works on an unknown number of nondeterministic choices, and it takes a 
function @k@ that works for every element in the collection. Here we present another
function @choices@ that uses delimited control operator @shift@ to capture that ``@k@''
when the program is direct-style and pass it to @nd@.

@@

The function @choices@ takes two arguments: a set of elements of type @T@, and an initial cache.
The return type of @choices@ is @(T, Cache)@ with an annotation \verb|@cps[Ans]|, which 
denotes that @choices@ will iteratively return an element of type @T@ from the collection
along with the latest cache once @f@ is invoked, but the final returned value of the 
function is type @Ans@.
The function @f@ is the delimited continuation captured by the @shift@ operator, which
is formed by the computation will be executed after the @choices@'s call-site.
Once the function @f@ (which is the function @k@ in @nd@) is invoked and returned, 
it returns to its call-site where @nd@ will be invoked on the rest elements of the collection.
Now we can use this function to sequentialize the @let@ case.

@@

For every nondeterministic fork-point in the abstract interpretation, we can simply call 
@choices@ on the set of possible choices and write the program sequentially as concrete 
interpreters, and the subsequent statements and expressions after the call-site become 
the delimited continuation @f@. We have two invocations of @choices@, which correspond 
to the two invocations of @nd@ in the refunctionalized form in the last section.
The function @choices@ also returns the latest cache to its left-hand side when invoked.
The cache will be still accumulated to the next call to @choices@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.

The function @analyze@ is also changed by calling the @reset@ operator around the function 
call of @aeval@ to set the delimiter.

>>>>

\begin{itemize}
  \item The components of the configuration are lifted as arguments of the evaluation
    function.
  \item The refunctionalized AAM collects all possible configurations, but our final target is an 
    abstract definitional interpreter that cares about final evaluation results.
    So starting from this section our abstract semantic artifact returns a set of final 
    values instead of collected configurations. 
  \item The configurations still play an important role in caching, where a cache
    is a mapping from configurations to sets of final values. The caching algorithm
    ensures termination of the analysis, as well as performs a sound over-approximation
    of all possible values and reachable paths.
  \item The function @aeval@ is simplified by introducing and using a nondeterministic 
    operator @nd@ which makes @aeval@ only uses one layer of continuation.
\end{itemize}

We will begin by discussing these changes separately, and finally see how they work together.

@@

We first transform the function @aeval@ to direct-style by removing the additional 
continuation argument @continue@ and sequentializing the expressions in @aeval@.
To enable using delimited control operators inside of @aeval@, we also add an 
annotation \verb|@cps[Ans]| on the return type, which tells the compiler to CPS
transform this function.
Back to direct-style on the case where @e@ appears as an atomic expression is straightforward: 
we simply unwrap the application of @continue@. Now the abstract evaluator @aeval@ looks as follows:

@@

The case where @e@ is a @let@ expression is not that obvious. As previously identified, 
function @nd@ works on an unknown number of nondeterministic choices, and it takes a 
function @k@ that works for every element in the collection. Here we present another
function @choices@ that uses delimited control operator @shift@ to capture that ``@k@''
when the program is written in direct-style and pass it to @nd@.

@@

The function @choices@ takes two arguments: a set of elements of type @T@, and an initial cache.
The return type of @choices@ is @(T, Cache)@ with an annotation \verb|@cps[Ans]|, which 
denotes that @choices@ will iteratively return an element of type @T@ from the collection
along with the latest cache once @f@ is invoked, but the final returned value of the 
function is type @Ans@.
The function @f@ is the delimited continuation captured by the @shift@ operator, which
is formed by the computation will be executed after the @choices@'s call-site.
Once the function @f@ (which is the function @k@ in @nd@) is invoked and returned, 
it returns to its call-site where @nd@ will be invoked on the remaining elements of the collection.
Now we can use this function to sequentialize expressions in the @let@ case.

@@

For every nondeterministic fork-point in the abstract interpretation, we can simply call 
@choices@ on the set of possible choices and write the program sequentially as concrete 
interpreters, and the subsequent statements and expressions after the call-site become 
the delimited continuation @f@. We have two applications of @choices@, which correspond 
to the two applications of @nd@ in the refunctionalized form in the last section.
The function @choices@ also returns the latest cache to its left-hand side when invoked.
The cache will be still accumulated to the following calls of @choices@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.

The function @analyze@ is also changed by calling the @reset@ operator around the function 
application of @aeval@ to set the delimiter.
